What We Eval in the Shadows

### 2021-08-11 paper/main.tex
Migrating \c to \code for normalized expression, environments, and origins of eval

\begin{tabular}{@{}p{.97\linewidth}}
  \vspace{1mm}
  \EE{$min(e)=\code{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
  those are constants such as integers, the remainder evaluate to a value. \Eval
  is needed when the expression denoting the values is constructed
  programmatically. Cases when a value is directly given to \eval, and \eval
  returns it unchanged, often occur when a computation has a default path and
  another, more interesting, path that requires evaluation. The following is a
  call site that always see a string; \packageNbCallSitesUniqueActualValue of
  sites only ever see a value. This shape usually run in few interpreter steps.
  The environment in which they evaluate is mostly irrelevant.
\begin{lstlisting}
  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
\vspace{2mm}
\EE{$min(e)=\code{X}$}\\[-2mm]\small Variable lookups are the most common shape,
\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
and variables normalize to \c X. A lookup is one step of execution, if a promise
is returned an arbitrary number of additional steps may be needed; the median is
\packageMinimizedmedianoperationsaRnd, suggesting that most variable read do
little work. Variables are often evaluated in environments that have been
constructed programmatically, \packageMinimizedpercentparentframesa of
expressions are evaluated in a function environment.
\vspace{2mm}
\EE{$min(e)=\code{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
\c{x[[24]]}. This takes a few more step of evaluation on average, and is
typically used in a function environment.

\vspace{2mm}
\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct
assignment {\tt <-} and parent environment assignment {\tt <\,\!<-} as well as
the \c{assign} function. It subsumes \c{\$}. Assignments represent the most
obvious source of side effects in \eval. Most cases originate from trivial code
generation where the assign expression is assembled using parse or substitute,
often in a loop. For example (taken from \c{PopED::get\_rse}), the following
code will populate the current scope with a number of variables:
%
\begin{lstlisting}
for (i in seq_along(args)) eval(parse(text=paste(args[i], "<-", i)))
\end{lstlisting}

\vspace{2mm}
\EE{$min(e)=\code{F()}$}\\[-2mm]\small Function calls without variable lookups or
assignment. A variable is allowed the function position, looking up
functions does not trigger computation unless the function's name
is bound to a promise. This shape typically does not perform much work.
\vspace{2mm}
\framebox{$min(e)=\code{F(X)}$}~\EE{$min(e)=\code{F(F(X))}$}\\[-2mm]\small Function
calls whose arguments may include variable references and nested calls. Together
they represent the most frequent shape and perform over 16 steps. They are
frequently evaluated in function environments.
\end{tabular}

\begin{tabular}{@{}p{.97\linewidth}}
  \medskip\EE{$min(e)=$ \code{FUN}}\\[-2mm]\small Function definitions without any
  computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
  \packageGeneralizedFunctionDefinitionSitesPercent of sites have function
  definitions nested in other expressions.

\vspace{2mm}
\EE{$min(e)=\code{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
they can be large we do not try to normalize their contents. They execute in a
median \packageMinimizedmedianoperationsjRnd steps, usually, in function
environments.

\vspace{2mm}
\EE{$min(e)=\code{model.frame}$}\\[-2mm]\small The \c{model.frame} function
returns a dataframe resulting from fitting the model described in a formula.
This shape subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the single most
popular function invoked from \eval. Each call does a median of 2K instructions.
\end{tabular}

@@

Most sites, \SESitesInOneClass, do all their side-effects consistently in one
kind of environments. The same happens at the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments.

@@

\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere,
making it \eval more dangerous than it already is. However, the data suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
  scenario, \eval in Javascript performs, store events can reach up to 40\% of
the events, and 7\% to 8\% of the \eval do side effects in the global scope. }
and that over half of them happen in a predictable environment. This gives a
ray of hope for a hypothetical R compiler. Even though \eval can do anything
anywhere, the data suggests most effects are sane.

@@

  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
    the case that the block is only passed to \eval rather than being directly
    part of in as in \lstinline|eval({...})|. 
\end{compactitem}

>>>>

\begin{tabular}{@{}p{.97\linewidth}}
  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
  those are constants such as integers, the remainder evaluate to a value. \Eval
  is needed when the expression denoting the values is constructed
  programmatically. Cases when a value is directly given to \eval, and \eval
  returns it unchanged, often occur when a computation has a default path and
  another, more interesting, path that requires evaluation. The following is a
  call site that always see a string; \packageNbCallSitesUniqueActualValue of
  sites only ever see a value. This shape usually run in few interpreter steps.
  The environment in which they evaluate is mostly irrelevant.
\begin{lstlisting}
  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
\\
\medskip\EE{$min(e)=$\code{X}}\\[-2mm]\small Variable lookups are the most common shape,
\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
and variables normalize to \c X. A lookup is one step of execution, if a promise
is returned an arbitrary number of additional steps may be needed; the median is
\packageMinimizedmedianoperationsaRnd, suggesting that most variable read do
little work. Variables are often evaluated in environments that have been
constructed programmatically, \packageMinimizedpercentparentframesa of
expressions are evaluated in a function environment.
\\
\medskip\EE{$min(e)=$\code{$}}\\[-2mm]\small This shape extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
\c{x[[24]]}. This takes a few more step of evaluation on average, and is
typically used in a function environment.
\\
\medskip\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct assignment {\tt
  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
function. It subsumes \c{\$}. Assignments
represent the most obvious source of side effects in \eval.
\\
\medskip\EE{$min(e)=$\code{F()}}\\[-2mm]\small Function calls without variable lookups or
assignment. A variable is allowed the function position, looking up
functions does not trigger computation unless the function's name
is bound to a promise. This shape typically does not perform much work.
\\
\medskip\framebox{$min(e)=$\code{F(X)}}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
calls whose arguments may include variable references and nested calls. Together
they represent the most frequent shape and perform over 16 steps. They are
frequently evaluated in function environments.
\end{tabular}

\begin{tabular}{@{}p{.97\linewidth}}
  \medskip\EE{$min(e)=$\code{FUN}}\\[-2mm]\small Function definitions without any
  computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
  \packageGeneralizedFunctionDefinitionSitesPercent of sites have function
  definitions nested in other expressions.
\\
\medskip\EE{$min(e)=$\code{BLOCK}}\\[-2mm]\small Multi-statement code blocks, as
they can be large we do not try to normalize their contents. They execute in a
median \packageMinimizedmedianoperationsjRnd steps, usually, in function
environments.
\\
\medskip\EE{$min(e)=$\code{model.frame}}\\[-2mm]\small The \c{model.frame} function
returns a dataframe resulting from fitting the model described in a formula.
This shape subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the single most
popular function invoked from \eval. Each call does a median of 2K instructions.
\end{tabular}

@@

Most sites, \SESitesInOneClass, do all their side-effects consistently in one
kind of environments. The same happens at the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for a hypothetical R
compiler. Even though \eval can do anything anywhere, the data suggests most
effects are sane.

@@

\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope. In
R, given the support for first-class environments, it can happen anywhere,
making it \eval more dangerous than it already is. However, the data suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
scenario, \eval in Javascript performs, store events can reach up to 40\% of the
events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
that over half of them happen in a predictable environment.

@@

  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
    the case that the block is only passed to \eval rather than being directly
    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
    fragment of a program to be evaluated in a particular way and in a particular
    environment. The latter is what makes it different from 0-argument closure.

    There is a number of use cases: unit testing frameworks (\eg \c{testthat},
    \c{testit}), code benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}),
    running code in parallel (\eg \c{foreach}, \c{doParallel}), or deferring
    code execution (\eg \c{withr}).

\end{compactitem}
